Data Science Applied to Ag - Final Project - ML

### 2025-04-25 code/Draft_script.qmd
d

@@

@@

@@

daymet_all_unnest

@@

#write_csv(daymet_all_unnest,
          #"../data/merged_fieldweatherdata.csv"
          )

@@

```{r}
#| message: false
#| warning: false

install.packages("ggridges")

@@

@@

fe_month <- fieldweather %>%
  # Selecting needed variables
  dplyr::select(year, site, hybrid, lon, lat, yday, 
                yield = adjusted_yield, 
                planting.date = planting_date, 
                harvest.data = harvest_date, 
                planting.doy = plant_doy, 
                harvest.doy = harvest_doy, 
                season.length = season_len, 
                soil.ph = soilpH, 
                soil.om.pct = om_pct, 
                soil.k.ppm = soilk_ppm, 
                soil.p.ppm = soilp_ppm, 
                dayl.s = dayl_s, #to rename variable name from "dayl_s" to dayl.s
                prcp.mm = prcp_mm_day, #to rename variable name to "prcp.mm"
                srad.wm2 = srad_w_m_2,#to rename variable name to "srad.wm2"
                tmax.c = tmax_deg_c, #to rename variable name to "tmax.c"
                tmin.c = tmin_deg_c,#to rename variable name to "tmin.c"
                vp.pa = vp_pa #to rename variable name to "vp.pa"
                ) %>%
  # Creating a date class variable  
  mutate(date_chr = paste0(year, "/", yday)) %>% 
  mutate(date = as.Date(date_chr, "%Y/%j")) %>% 
  # Extracting month from date  
  mutate(month = month(date)) %>% 
  mutate(month_abb = month(date, label = T)) #To get abbreviated month name e.g., Jan, Feb, Mar,...,Dec #month_abb is in "ord" (ordinal) format

@@

Now, let's summarize daily weather variables based on month.

@@

fe_month_sum <- fe_month %>%
  group_by(year, site, hybrid, month_abb, yield) %>% #If we do a summarise() after group_by(), any column that's not in the group_by() is gone, so we need to include "strength_gtex" in the group_by() to include it in the data frame because "strength_gtex" is our response variable so we must keep it in the data frame 
  #Because we are gonna be applying a "summarize()" function to different columns, we are gonna use a function called "across()" 
  summarise(across(.cols = c(soil.ph, 
                             soil.om.pct, 
                             soil.k.ppm, 
                             soil.p.ppm,
                             dayl.s,
                             srad.wm2,
                             tmax.c,
                             tmin.c,
                             vp.pa),
                   .fns = mean, #do not indicate the actual function "mean()", just use the word "mean"
                   .names = "mean_{.col}"), #specifying the weather variables that we want their mean as new column variables #1st across() is applying the "mean" function (to summarize "mean")
            across(.cols = prcp.mm,
                   .fns = sum,
                   .names = "sum_{.col}"
                   ) #specifying the weather variable (prcp.mm) that we want its sum as new column variable #2nd across() is summarizing sum #2nd across() is applying the "sum" function (to summarize "sum")
            ) %>%
  ungroup() #To convert from "group" to "tibble"

@@

Let's check tmax.c and prcp.mm for the first site-year and month.

[Note: the following code chunk is for double checking to make sure that we did everything okay the way we intended]

@@

fe_month_sum_wide <- fe_month_sum %>%
  pivot_longer(cols = mean_soil.ph:sum_prcp.mm) %>% 
  mutate(varname = paste0(name, "_", month_abb)) %>% 
  dplyr::select(-name, -month_abb) %>% 
  pivot_wider(names_from = varname,
              values_from = value) %>%
  # Rounding to one decimal point
  mutate(across(5:ncol(.), ~round(., 1) )) %>%
  select(-mean_soil.ph_NA, -mean_soil.om.pct_NA, -mean_soil.k.ppm_NA, -mean_soil.p.ppm_NA, -mean_dayl.s_NA, -mean_srad.wm2_NA, -mean_tmax.c_NA, -mean_tmin.c_NA, -mean_vp.pa_NA, -sum_prcp.mm_NA)

@@

>>>>

weather_daymet_all

@@

View(weather_daymet_unnest)

@@

The following code chunk will merge the weather data retrieved from Daymet to the "merged_clean" data

@@

fieldweather <- daymet_all_unnest

fieldweather

@@

#write_csv(daymet_all_unnest,
          #"../data/merged_fieldweatherdata.csv"
          #)

@@

#install.packages("ggridges")

@@

The following code chunk will keep desired variables that we will use further and get abbreviated month name based on the date.

@@

fe_month <- fieldweather %>%
  # Selecting needed variables
  dplyr::select(year, 
                site, 
                hybrid, 
                lon, 
                lat, 
                yday, 
                yield = adjusted_yield, 
                prev_crop,
                plant_doy,
                harvest_doy,
                soil.ph = soilpH, 
                soil.om.pct = om_pct, 
                soil.k.ppm = soilk_ppm, 
                soil.p.ppm = soilp_ppm, 
                dayl.s = dayl_s, #to rename variable name from "dayl_s" to dayl.s
                prcp.mm = prcp_mm_day, #to rename variable name to "prcp.mm"
                srad.wm2 = srad_w_m_2,#to rename variable name to "srad.wm2"
                tmax.c = tmax_deg_c, #to rename variable name to "tmax.c"
                tmin.c = tmin_deg_c,#to rename variable name to "tmin.c"
                vp.pa = vp_pa #to rename variable name to "vp.pa"
                ) %>%
  # Creating a date class variable  
  mutate(date_chr = paste0(year, "/", yday)) %>% 
  mutate(date = as.Date(date_chr, "%Y/%j")) %>% 
  # Extracting month from date  
  mutate(month = month(date)) %>% 
  mutate(month_abb = month(date, label = T)) #To get abbreviated month name e.g., Jan, Feb, Mar,...,Dec #month_abb is in "ord" (ordinal) format

@@

The following code chunk will feature engineer planting and harvest stages based on the following cut-offs:

Planting
•	early: DOY ≤ 120
•	normal: 121–150
•	late: > 150

Harvest
•	early: DOY ≤ 275
•	normal: 276–305
•	late: > 305

fe_month <- fe_month%>%
  mutate(
    planting_stages = case_when(
      plant_doy <= 120                     ~ "early",
      plant_doy >= 121 & plant_doy <= 150  ~ "normal",
      plant_doy > 150                      ~ "late"
    ) %>% 
      fct_relevel("early","normal","late"),

    harvest_stages = case_when(
      harvest_doy <= 275                    ~ "early",
      harvest_doy >= 276 & harvest_doy <= 305 ~ "normal",
      harvest_doy > 305                     ~ "late"
    ) %>% 
      fct_relevel("early","normal","late")
  ) %>%
  select(-plant_doy, -plant_doy)

fe_month

The following code chunk will summarize daily weather variables based on month.

@@

fe_month_sum <- fe_month %>%
  group_by(year, site, hybrid, lon, lat, month_abb, yield, prev_crop, planting_stages, harvest_stages) %>% #If we do a summarise() after group_by(), any column that's not in the group_by() is gone, so we need to include "strength_gtex" in the group_by() to include it in the data frame because "strength_gtex" is our response variable so we must keep it in the data frame 
  #Because we are gonna be applying a "summarize()" function to different columns, we are gonna use a function called "across()" 
  summarise(across(.cols = c(soil.ph, 
                             soil.om.pct, 
                             soil.k.ppm, 
                             soil.p.ppm,
                             dayl.s,
                             srad.wm2,
                             tmax.c,
                             tmin.c,
                             vp.pa),
                   .fns = mean, #do not indicate the actual function "mean()", just use the word "mean"
                   .names = "mean_{.col}"), #specifying the weather variables that we want their mean as new column variables #1st across() is applying the "mean" function (to summarize "mean")
            across(.cols = prcp.mm,
                   .fns = sum,
                   .names = "sum_{.col}"
                   ) #specifying the weather variable (prcp.mm) that we want its sum as new column variable #2nd across() is summarizing sum #2nd across() is applying the "sum" function (to summarize "sum")
            ) %>%
  ungroup() #To convert from "group" to "tibble"

@@

The following code chunk will check "tmax.c" and "prcp.mm" for the first site-year and month for the purpose of double checking to make sure that we did everything okay the way we intended.

@@

fe_month_sum_wide <- fe_month_sum %>%
  pivot_longer(cols = mean_soil.ph:sum_prcp.mm) %>% 
  mutate(varname = paste0(name, "_", month_abb)) %>% 
  dplyr::select(-name, -month_abb) %>% 
  pivot_wider(names_from = varname,
              values_from = value) %>%
  # Rounding to one decimal point
  mutate(across(10:ncol(.), ~round(., 1) )) %>%
  select(-mean_soil.ph_NA, -mean_soil.om.pct_NA, -mean_soil.k.ppm_NA, -mean_soil.p.ppm_NA, -mean_dayl.s_NA, -mean_srad.wm2_NA, -mean_tmax.c_NA, -mean_tmin.c_NA, -mean_vp.pa_NA, -sum_prcp.mm_NA)

@@

View(fe_month_sum_wide)
